Viscous Transport in Eroding Porous Media

### 2019-10-10 docs/transport/erosion_transport.tex
finished up formatting and looked at Nick's changes

\begin{abstract} 
  Transport of viscous fluid through porous media is a direct
  consequence of the pore structure. Here we investigate transport
  through \nick{a specific class of two-dimensional porous geometries, 
  namely those formed by fluid-mechanical erosion.}  
  We investigate the tortuosity and
  dispersion by analyzing the first two statistical moments of tracer
  trajectories. \nick{For most initial configurations, tortuosity decreases in time 
  as a result of erosion increasing the porosity. 
  However, we find that tortuosity can also increase transiently in certain cases.}
  \nick{The porosity-tortuosity relationships that result from our simulations 
  are compared against models available in the literature.}
  \nick{Asymptotic dispersion rates are also strongly affected by the
  erosion process, as well as by the number and
  distribution of the eroding bodies.}  
  Finally, we analyze the pore size
  distribution of an eroding geometry. The simulations are performed by
  combining a high-fidelity boundary integral equation solver for the
  fluid equations, a second-order stable time stepping method to
  simulate erosion, and new numerical methods to stably and accurately
  resolve nearly-touching eroded bodies and particle trajectories near
  the eroding bodies.

@@

This study consists of two main undertakings: first, high-fidelity
simulations of eroding porous media, and, second, characterization of
tracer transport through the resulting eroded geometries. Our modeling
efforts build on previous work~\citep{ris-moo-chi-she-zha2012,
moore2013self, moo2017}, in particular recent numerical methods
developed to simulate erosion in the Stokes-flow
regime~\citep{qua-moo2018}. We, however, make key improvements to the
numerical methods \nick{(I killed the reference here because I think its redundant)} to enable simulations of more
realistic, dense suspensions of eroding grains
(figures~\ref{fig:Eroding50vel} and~\ref{fig:Eroding50vort}).  Then, to
characterize transport through these configurations, we examine
coarse-grained variables through statistical analysis of tracer
trajectories.

@@

Owing to the scales present in groundwater flow~\citep{bea1972}, we
model the hydrodynamics with the two-dimensional incompressible Stokes
equations.  Meanwhile, individual grains erode at a rate proportional to
the hydrodynamic shear stress~\citep{wan-fel2004,
ris-moo-chi-she-zha2012, moore2013self, par-izu2000}.  \nick{We compute
the vorticity in the fluid bulk since, on solid boundaries, vorticity
reduces to shear and thus provides a convenient way to simultaneously
visualize local erosion rates and changes in the surrounding flow}
(figure~\ref{fig:Eroding50vort}).  A variety of numerical methods for
the fluid solver are available including lattice-gas or lattice
Boltzmann~\citep{kop-kat-tim1996, dar-mcc1998}, finite
volumes~\citep{suo-liu-gan2019, den-ica-hid2018,
sie-ili-pri-riv-gua2019}, finite differences~\citep{leb-ded-dav-bou2007,
knu-car2005}, and cellular automata~\citep{rot1988}.  However,
simulating erosion requires a method that is high-order, can resolve
nearly-touching arbitrarily shaped grains, and can accurately compute
derivatives of the velocity field.  Since the governing equations are
linear and homogeneous, we address these challenges by converting the
fluid equations to a boundary integral equation (BIE).

@@

Given the discretization points of $\bd\Omega$, the trapezoid rule
results in the collocation method for~\eqref{eqn:BIE}
\todo[inline]{left off here with formatting equations}
\begin{subequations}
\label{eqn:trapLinearSystem}
  \begin{alignat}{3}
  \UU_\ell &= \sum_{j=1}^{N_\out} 
    w^0_j D(\yy^0_\ell,\yy^0_j) \eeta_j^0 +
  \sum_{k=1}^M \sum_{j=1}^{N_\iin}
    w^k_j D(\yy^0_\ell,\yy^k_j) \eeta^k_j +
  \sum_{j=1}^{N_\out} w^0_j N_0(\yy^0_\ell,\yy^0_j)\eeta_j^0 
    \nonumber \\
  &\quad+\sum_{k=1}^M S[\llambda_k,\cc_k](\yy^0_\ell) + 
  \sum_{k=1}^M R[\xi_k,\cc_k](\yy^0_\ell),
    &&\hspace{-60pt} \ell = 1,\ldots,N_\out, \\
%
  \mathbf{0} &= \sum_{j=1}^{N_\out} 
    w^0_j D(\yy^m_\ell,\yy^0_j) \eeta_j^0 +
  \sum_{k=1}^M \sum_{j=1}^{N_\iin}
    w^k_j D(\yy^m_\ell,\yy^k_j) \eeta^k_j \nonumber \\
  &\quad+\sum_{k=1}^M S[\llambda_k,\cc_k](\yy^m_\ell) + 
  \sum_{k=1}^M R[\xi_k,\cc_k](\yy^m_\ell),
    &&\hspace{-60pt}\ell = 1,\ldots,N_\iin, \: m=1,\ldots,M, \\
%
  \llambda_m &= \frac{1}{2\upi} \sum_{j=1}^{N_\iin} w_j^m \eeta^m_j,
  \quad \xi_m = \frac{1}{2\upi} \sum_{j=1}^{N_\iin} w_j^m
    (\yy^m_j - \cc_m)^\perp \cdot \eeta^m_j,
    &&\hspace{-60pt} m=1,\ldots,M,
\end{alignat}
\end{subequations}
where $w^k_j$ are quadrature weights that depend on $N_\iin$, $N_\out$,
and the lengths of $\gamma^k$ and $\Gamma$, and $D$ is the kernel of the
Stokes double-layer potential defined in
equation~\eqref{eqn:velocityDLP}.  Since the kernel is smooth, the
diagonal terms $D(\yy_j^m,\yy_j^m)$ are replaced with the appropriate
curvature-dependent limiting term.  The linear
system~\eqref{eqn:trapLinearSystem} is a well-conditioned second-kind
integral equation that is solved iteratively with GMRES.  If the number
of discretization points is sufficiently large, then the solution
of~\eqref{eqn:trapLinearSystem} is an accurate approximation of the
density function, Stokeslets, and rotlets.   Then, if $\xx \in \Omega$
is sufficiently far from $\bd\Omega$, an approximation of the velocity,
deformation tensor, and vorticity due to the double-layer potential is
found by applying the trapezoid rule to \eqref{eqn:velocityDLP},
\eqref{eqn:deformationDLP}, and~\eqref{eqn:vorticityDLP}, respectively.
In particular,
\begin{align}
  \label{eqn:trap}
  \uu(\xx) &= \sum_{j=1}^{N_\out} w^0_j D(\xx,\yy^0_j) \eeta_j +
  \sum_{k=1}^M \sum_{j=1}^{N_\iin} w^k_j D(\xx,\yy^k_j) \eeta^k_j, \\
%
  \ssigma(\xx) &= \sum_{j=1}^{N_\out} w^0_j \mathbf{D}_\ssigma(\xx,\yy^0_j) \eeta_j +
  \sum_{k=1}^M \sum_{j=1}^{N_\iin} w^k_j \mathbf{D}_\ssigma(\xx,\yy^k_j)
  \eeta^k_j, \\
%
  w(\xx) &= \sum_{j=1}^{N_\out} w^0_j D_\omega(\xx,\yy^0_j) \eeta_j +
  \sum_{k=1}^M \sum_{j=1}^{N_\iin} w^k_j D_\omega(\xx,\yy^k_j)
  \eeta^k_j.
\end{align}
The contributions due to the Stokeslets and rotlets require no
quadrature and are easily included in the velocity, deformation tensor,
and vorticity.  Finally, to compute the shear stress, we apply the
trapezoid rule with Fourier differentiation to approximate the jump
term~\eqref{eqn:deformationJump} of the deformation tensor, and then the
tensor is applied to the normal and tangent vectors as defined in
equation~\eqref{eqn:shearStress}.

>>>>

\begin{abstract} 
  Transport of viscous fluid through porous media is a direct
  consequence of the pore structure. Here we investigate transport
  through a specific class of two-dimensional porous geometries, namely
  those formed by fluid-mechanical erosion.  We investigate the
  tortuosity and dispersion by analyzing the first two statistical
  moments of tracer trajectories. For most initial configurations,
  tortuosity decreases in time as a result of erosion increasing the
  porosity.  However, we find that tortuosity can also increase
  transiently in certain cases.  The porosity-tortuosity relationships
  that result from our simulations are compared against models available
  in the literature.  Asymptotic dispersion rates are also strongly
  affected by the erosion process, as well as by the number and
  distribution of the eroding bodies. Finally, we analyze the pore size
  distribution of an eroding geometry. The simulations are performed by
  combining a high-fidelity boundary integral equation solver for the
  fluid equations, a second-order stable time stepping method to
  simulate erosion, and new numerical methods to stably and accurately
  resolve nearly-touching eroded bodies and particle trajectories near
  the eroding bodies.

@@

This study consists of two main undertakings: first, high-fidelity
simulations of eroding porous media, and, second, characterization of
tracer transport through the resulting eroded geometries. Our modeling
efforts build on previous work~\citep{ris-moo-chi-she-zha2012,
moore2013self, moo2017}, in particular recent numerical methods
developed to simulate erosion in the Stokes-flow
regime~\citep{qua-moo2018}. We, however, make key improvements to the
numerical methods to enable simulations of more realistic, dense
suspensions of eroding grains (figures~\ref{fig:Eroding50vel}
and~\ref{fig:Eroding50vort}).  Then, to characterize transport through
these configurations, we examine coarse-grained variables through
statistical analysis of tracer trajectories.

@@

Owing to the scales present in groundwater flow~\citep{bea1972}, we
model the hydrodynamics with the two-dimensional incompressible Stokes
equations.  Meanwhile, individual grains erode at a rate proportional to
the hydrodynamic shear stress~\citep{wan-fel2004,
ris-moo-chi-she-zha2012, moore2013self, par-izu2000}.  We compute the
vorticity in the fluid bulk since, on solid boundaries, vorticity
reduces to shear and thus provides a convenient way to simultaneously
visualize local erosion rates and changes in the surrounding flow
(figure~\ref{fig:Eroding50vort}).  A variety of numerical methods for
the fluid solver are available including lattice-gas or lattice
Boltzmann~\citep{kop-kat-tim1996, dar-mcc1998}, finite
volumes~\citep{suo-liu-gan2019, den-ica-hid2018,
sie-ili-pri-riv-gua2019}, finite differences~\citep{leb-ded-dav-bou2007,
knu-car2005}, and cellular automata~\citep{rot1988}.  However,
simulating erosion requires a method that is high-order, can resolve
nearly-touching arbitrarily shaped grains, and can accurately compute
derivatives of the velocity field.  Since the governing equations are
linear and homogeneous, we address these challenges by converting the
fluid equations to a boundary integral equation (BIE).

@@

Given the discretization points of $\bd\Omega$, the trapezoid rule
results in the collocation method for~\eqref{eqn:BIE}
\begin{subequations}
\label{eqn:trapLinearSystem}
  \begin{alignat}{3}
  \UU_\ell &= \sum_{j=1}^{N_\out} 
    w^0_j D(\yy^0_\ell,\yy^0_j) \eeta_j^0 +
  \sum_{k=1}^M \sum_{j=1}^{N_\iin}
    w^k_j D(\yy^0_\ell,\yy^k_j) \eeta^k_j +
  \sum_{j=1}^{N_\out} w^0_j N_0(\yy^0_\ell,\yy^0_j)\eeta_j^0 
    \nonumber \\
  &+\sum_{k=1}^M S[\llambda_k,\cc_k](\yy^0_\ell) + 
  \sum_{k=1}^M R[\xi_k,\cc_k](\yy^0_\ell), 
  &&\hspace{-124pt}\ell = 1,\ldots,N_\out, \\
%
  \mathbf{0} &= \sum_{j=1}^{N_\out} 
    w^0_j D(\yy^m_\ell,\yy^0_j) \eeta_j^0 +
  \sum_{k=1}^M \sum_{j=1}^{N_\iin}
    w^k_j D(\yy^m_\ell,\yy^k_j) \eeta^k_j \nonumber \\
  &+\sum_{k=1}^M S[\llambda_k,\cc_k](\yy^m_\ell) + 
  \sum_{k=1}^M R[\xi_k,\cc_k](\yy^m_\ell),
    &&\hspace{-124pt}m=1,\ldots,M, \: \ell = 1,\ldots,N_\iin,  \\
%
  \llambda_m &= \frac{1}{2\upi} \sum_{j=1}^{N_\iin} 
    w_j^m \eeta^m_j, 
  &&\hspace{-124pt}m = 1,\ldots,M \\ 
  \xi_m &= \frac{1}{2\upi} \sum_{j=1}^{N_\iin} 
    w_j^m (\yy^m_j - \cc_m)^\perp \cdot \eeta^m_j,
  &&\hspace{-124pt}m=1,\ldots,M,
\end{alignat}
\end{subequations}
where $w^k_j$ are quadrature weights that depend on $N_\iin$, $N_\out$,
and the lengths of $\gamma^k$ and $\Gamma$, and $D$ is the kernel of the
Stokes double-layer potential defined in
equation~\eqref{eqn:velocityDLP}.  Since the kernel is smooth, the
diagonal terms $D(\yy_j^m,\yy_j^m)$ are replaced with the appropriate
curvature-dependent limiting term.  The linear
system~\eqref{eqn:trapLinearSystem} is a well-conditioned second-kind
integral equation that is solved iteratively with GMRES.  If the number
of discretization points is sufficiently large, then the solution
of~\eqref{eqn:trapLinearSystem} is an accurate approximation of the
density function, Stokeslets, and rotlets.   Then, if $\xx \in \Omega$
is sufficiently far from $\bd\Omega$, an approximation of the velocity,
deformation tensor, and vorticity due to the double-layer potential is
found by applying the trapezoid rule to \eqref{eqn:velocityDLP},
\eqref{eqn:deformationDLP}, and~\eqref{eqn:vorticityDLP}, respectively.
In particular,
\begin{align}
  \label{eqn:trap}
  \uu(\xx) &= \sum_{j=1}^{N_\out} w^0_j D(\xx,\yy^0_j) \eeta_j +
  \sum_{k=1}^M \sum_{j=1}^{N_\iin} w^k_j D(\xx,\yy^k_j) \eeta^k_j, \\
%
  \ssigma(\xx) &= \sum_{j=1}^{N_\out} w^0_j \mathbf{D}_\ssigma(\xx,\yy^0_j) \eeta_j +
  \sum_{k=1}^M \sum_{j=1}^{N_\iin} w^k_j \mathbf{D}_\ssigma(\xx,\yy^k_j)
  \eeta^k_j, \\
%
  w(\xx) &= \sum_{j=1}^{N_\out} w^0_j D_\omega(\xx,\yy^0_j) \eeta_j +
  \sum_{k=1}^M \sum_{j=1}^{N_\iin} w^k_j D_\omega(\xx,\yy^k_j)
  \eeta^k_j.
\end{align}
The contributions due to the Stokeslets and rotlets require no
quadrature and are easily included in the velocity, deformation tensor,
and vorticity.  Finally, to compute the shear stress, we apply the
trapezoid rule with Fourier differentiation to approximate the jump
term~\eqref{eqn:deformationJump} of the deformation tensor, and then the
tensor is applied to the normal and tangent vectors as defined in
equation~\eqref{eqn:shearStress}.
